---
title: "Arrays"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(echo = TRUE)
library(pacman)
p_load(tidyverse)
p_load(nycflights13)
p_load(ggplot2)
p_load(Deriv)
p_load(deSolve)
p_load(phaseR)
p_load(gridExtra)
theme_set(theme_minimal())
```

## Arrays

### Define arrays and matrices

- An array can have any number of dimensions and requires a dim parameter
- A matrix has always 2 dimensions and requires a nrow  parameter
- You can specify row and col names througgh dimnames 
- It is also to specity a one dimensional array, **but be careful with it because it becomes a column, not a row** 

```{r}

# A 3-d array with names
three_d_array <- array(
  1:24,
  dim = c(4,3,2),
  dimnames = list(
    c("one", "two", "three", "four"),
    c("ein", "zwei", "drie"),
    c("un", "deux")
  )
)

# A 2-d array with names
two_d_array <- array(
  1:12,
  dim = c(4,3),
  dimnames = list(
    c("one", "two", "three", "four"),
    c("ein", "zwei", "drie")
  )
)
two_d_array

one_d_array <- array(
  1:4,
  dim = c(4),
  dimnames = list(
    c("one", "two", "three", "four")
  )
)
one_d_array

# An identical matrix 
a_matrix <- matrix(
  1:12,
  nrow  = 4,
  dimnames = list(
    c("one", "two", "three", "four"),
    c("ein", "zwei", "drie")
  )
)
a_matrix 

# The same matrix without names 
a_matrix_nn <- matrix(
  1:12,
  nrow  = 4,
) 
a_matrix_nn
```


### Getting info on matrices and arrays

- Dimensions of an array or matrix can be shown with dim
- For matrices you can also use nrow and ncol
- For arrays nrow and ncol can only show the first two dimensions 

```{r}

dim(three_d_array)
dim(two_d_array)
dim(one_d_array)
dim(a_matrix)

# As expected this works
nrow(a_matrix)
ncol(a_matrix)

# This works also, but clearly you cannot get the first two dimensions
nrow(three_d_array)
ncol(three_d_array)

nrow(one_d_array)
ncol(one_d_array)


```

### Names of matrices and arrays

- For matrices you can also use rownames and colnames
- You can delete them or change them

```{r}
dimnames(three_d_array)
dimnames(a_matrix)

dimnames(three_d_array)[[1]]
dimnames(three_d_array)[[2]] 
dimnames(three_d_array)[[3]] 

rownames(a_matrix)
colnames(a_matrix)

```


### Changing row and col names

- Names can be inspected, changed and deleted with dimnames
- For matrices you use rownames and colnames

```{r}

dimnames(three_d_array)[[2]] <- c("G", "H", "I")
three_d_array

# dimnames(three_d_array)[[1]] <- c(NULL, NULL, NULL, NULL)
three_d_array

# Change rownames
rownames(a_matrix) <- c("one_1", "two_2", "three_3", "four_4")
a_matrix

# Delete rownames
rownames(a_matrix) <- NULL
a_matrix
```



### Subsetting arrays
```{r}

a_matrix <- matrix(
  1:12,
  nrow  = 4,
  dimnames = list(
    c("one", "two", "three", "four"),
    c("ein", "zwei", "drie")
  )
)

a_matrix 

# Select two rows and two cols
a_matrix[c("one", "two"), c(1,3)]

# Select two rows and one col
s2 <- a_matrix[c("one", "two"), -c(1,3)]
s2
nrow(s2)
ncol(s2)

# Select one row and two cols
s3 <- a_matrix[c("one"), c(1,3)]
s3
nrow(s3)
ncol(s3)
```

### Accessing matrix elements

- For a matrix you need to specify two dimension
- If you specify one dimension, the elements are assumed to be in a sequence (column first)


```{r}
a_matrix  <- matrix(seq(23,34), nrow = 3)
a_matrix

# First row
a_matrix[1,]

# Second column
a_matrix[,2]

# An element
a_matrix[1,2]

# Peculiar, here it interprets the matris as a column vector
a_matrix[8]

```


