---
title: "Lists"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Lists


### Creating lists
- Unlike vectors, lists can contain disimilar elements

```{r}
a_list <- list(
  list (1, 1, 2, 5, 14, 42),
  month.abb,
  matrix(c(3, -8, 1, -3), nrow = 2),
  asin
)
a_list

b_list <- list(
  catalan = list (1, 1, 2, 5, 14, 42),
  months = month.abb,
  involuntary = matrix(c(3, -8, 1, -3), nrow = 2),
  arcsin = asin
)
b_list

names(a_list) <- c("new name 1", "new name 2", "new name 3", "new name 4")
a_list
```


### Information on lists

- dim, nrow, ncol do not work on lists

```{r}
str(a_list)
class(a_list)
dim(a_list)
nrow(a_list)
ncol(a_list)
```

### Accessing list elements


- Using single brackets always returns a list
- Using double brackets returns the element contained within

#### Single and double brackets
```{r}

# Difference between single and double brackets
a_list[1]
a_list[[1]]
class(a_list[1])
class(a_list[[1]])

a_list[2]
a_list[[2]]
class(a_list[2])
class(a_list[[2]])

a_list[3]
a_list[[3]]
class(a_list[3])
class(a_list[[3]])

a_list[4]
a_list[[4]]
class(a_list[4])
class(a_list[[4]])
```

#### Indexing in list

```{r}

# Element 3 from list element 2 is of class character 
a_list[[2]]
a_list[[2]][3]
class(a_list[[2]][3])

# Element 3 from list element 2 is of list character 
a_list[[1]]
a_list[[1]][6]
class(a_list[[1]][6])

# Element 3 from list is of matrix  character 
a_list[[3]]
class(a_list[[3]])

# The first row of that matrix 
a_list[[3]][1,]
class(a_list[[3]][1,])

# The second col of that matrix 
a_list[[3]][,2]

# Retrieving multiple elements from a list by index
a_list[c(1,3)]

# Retrieving elements from a list by name
a_list["new name 2"]
a_list[c("new name 4","new name 2")]

# Note you can eve retrievce in duplicate
a_list[c("new name 2","new name 1", "new name 2", "new name 2")]
```
### Removing elements in a list

```{r}
c_list <- list("Element 1","Element 2", "Element 3", "Element 4")
c_list

c_list[c(1,3)] <- NULL
c_list

```

### Setting a list elements to NULL

- The point to note here is to use list(NULL) and not just NULL

```{r}
c_list <- list("Element 1","Element 2", "Element 3", "Element 4")
c_list

c_list[c(1,3)] <- list(NULL)
c_list

```

