---
title: "Recipes"
author: "Hans"
date: "9/11/2019"
output:
  html_document: default
  word_document: default
---


Important references:

http://www.cookbook-r.com

https://www.r-graph-gallery.com




# Setup - always run

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pacman)
p_load(tidyverse)
p_load(nycflights13)
p_load(ggplot2)
p_load(Deriv)
p_load(deSolve)
p_load(phaseR)
p_load(gridExtra)
theme_set(theme_minimal())
```



# Control structures

## if elseif else
```{r}

i = 2

if ( i > 0) {
  print ("i > 0") 
} else if (i < 0) {
  print ("i < 0") 
} else {
  print ("i = 0") 
}
```


## For loop
```{r}

# Select i from a vector
for (i in c(1, 5, 12)) {
  cat("i =", i, "\n")
}

# Select i from a range
for(i in 1:4) {
  cat("i =", i, "\n")
}

# Select i from a range - note that counting backward also  works
for(i in 4:1) {
  cat("i =", i, "\n")
}
```

## While loop
```{r}
i <- 0
while (i < 10) {
  cat ("i = ", i, "\n")
  i <- i + 1
}

```


## Repeat loop
```{r}

i <- 0
repeat {
  cat ("i = ", i, "\n")
  i <- i + 1
  if (i > 10) {
    break
  }
}
```


# Get info on functions and variables
```{r}

# Get all variables defined
ls ()

# Get particular variables defined
ls (pattern = "nu")

# Get all functions starting with d
apropos("^d")

# Get all functions ending with d
apropos("d$")

# Get all functions that have a number in the name
apropos("[0-9]")

# Get all functions that have tye string "vector" in the name
apropos("vector")

# Get the class of a variuble
class("string")

# Get the structure of a variable
num <- runif(30)
fac <- factor(sample(letters[1:5], 30, replace = TRUE))
bool <- sample(c(TRUE, FALSE), 30, replace = TRUE)
df <- data.frame(num, fac, bool)

str(num)
str(fac)
str(df)

# Get a summary of a variable
summary(num)
summary(fac)
summary(df)

attributes(num)
attributes(fac)
attributes(df)
```

# Operators
```{r}

# Normal division
10 / 3

# Integer division
10 %/% 7

# Remainder
10 %% 7

# Checking equality
2 == 2

# Checking float eqiuality
sqrt(2)^2 == 2
all.equal(sqrt(2)^2, 2)
```



# Mathematics



## Get derivative and plot
```{r}

f1 <- function(x, a, b, c, h) {
  a*x^2 + b*exp(x) + c - h
}

df.f1 = Deriv(f1, "x")
df.f1



a_val = 2
b_val = 3
c_val = 5
h_val = 0

# Define vector of x values and corresponding function values for plotting
xmin  = -10
xmax  = 3
xstep = 0.01

xvals  = seq(xmin, xmax, xstep) 
yvals  = f1(xvals, a_val, b_val, c_val, h_val)
dyvals = df.f1(xvals, a_val, b_val, c_val, h_val)

df = data.frame(xvals, yvals, dyvals)

# Plot the function
ggplot(df) +
  geom_line(mapping = aes(x = xvals, y = yvals), color = "blue") +
  labs(title = "Plot Title", 
       x = "Horizontal label", 
       y = "Vertical label")

# Plot the derivative
ggplot(df) +
  geom_line(mapping = aes(x = xvals, y = dyvals), color = "blue") +
  labs(title = "Plot Title", 
       x = "Horizontal label", 
       y = "Vertical label")

# Or plot them in one graph
ggplot(df) +
  geom_line(mapping = aes(x = xvals, y = yvals), color = "blue") +
  geom_line(mapping = aes(x = xvals, y = dyvals), color = "red") +
  labs(title = "Plot Title", 
       x = "Horizontal label", 
       y = "Vertical label")

```

## Numerically integrate
```{r }

myfun <- function(x, a) {
  1 / (1 + exp(-x + 10) / a)
}

myfun <- function(x, a) {
  a * x
}

(val = integrate(myfun, 0, 10, a = 2))

xseq = seq(0, 10, 0.01)
yseq = myfun(xseq, 2)
df   = data.frame(xseq, yseq)

ggplot(df) +
  geom_line(mapping = aes(x = xseq, y = yseq), color = "blue") +
  labs(title = "Plot Title", 
       x = "Horizontal label", 
       y = "Vertical label")


```

## Solve equation
```{r}

f1 <- function(x) {
  2*x^2 - 5
}

# Plot the function, so you cn select x values witj positive and negative y value
xseq = seq(0, 3, 0.01)
yseq = f1(xseq)
df = data.frame(xseq, yseq)

ggplot(df) +
  geom_line(mapping = aes(x = xseq, y = yseq), color = "blue") +
  labs(title = "Plot Title", 
       x = "Horizontal label", 
       y = "Vertical label")

xcross = uniroot(f1, lower = 1, upper = 2, tol = 0.000001)
xcross$root
(f1(xcross$root))

```

# ODE

## Solve simple ODE 
```{r}

ode_func = function(timepoint, state, parameters) { 
  with (as.list(c(state, parameters)), {
    dY = a * Y  # Function definition
    list(c(dY)) 
  })
}

pars = c(a = 0.2)
inistate = c(Y = 1)
time_series = seq(0, 10, by = 0.1)

out = ode(y = inistate, 
          times = time_series, 
          func = ode_func, 
          parms = pars)

df = data.frame(out)
ggplot(data = df) +
  geom_line(mapping = aes(x=time, y = Y))

```


## Phase portrait and stability
```{r}

# In this function the paremeter is Y
# This needs to be entered in state.names

ode_func = function(timepoint, state, parameters) { 
  with (as.list(c(state, parameters)), {
    dY = a * Y  # Function definition
    list(c(dY)) 
  })
}

pars <- c(a = -0.2)

phaseport = phasePortrait(ode_func, 
                          parameters = pars, 
                          ylim = c(-2, 2), 
                          points = 10, 
                          xlab = "Y", 
                          ylab = "dY/dt (a = 1)", 
                          state.names = c("Y")     
                          ) 
grid() 

stab.expgrowth = stability(ode_func, 
                           ystar = 0, 
                           parameters = pars, 
                           system = "one.dim", 
                           state.names = c("Y")    # Do not forget this
                           )

```

# Simple ggplot recipes

Refer to:https://www.r-graph-gallery.com


## Scatterplot
```{r}

# Take mpg as data set

ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = as.factor(cyl)))
             
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = as.factor(cyl)), 
             alpha = 0.3,
             size = 1,
             shape = 25) +  # Values from 0..25
  labs(color = "Cylinder cont.") +
  labs(x = "New x label") +
  labs(y = "New y label") +
  labs(title = "New Plot", subtitle = "New subtitle") +
  labs(caption = "(based on data from ...)")


```


## Line plot
```{r}

# Example 1

xValue <- 1:100
yValue <- cumsum(rnorm(100))
dataf  <- data.frame(xValue, yValue)

ggplot(data = dataf) +
  geom_line(mapping = aes(x = xValue, y = yValue), 
            alpha = 0.3,
            size = 1) +
  labs(color = "Cylinder cont.") +
  labs(x = "New x label") +
  labs(y = "New y label") +
  labs(title = "New Plot", subtitle = "New subtitle") +
  labs(caption = "(based on data from ...)" )

# Example 2

x <- seq(-4,4,0.1)
y <- dnorm(x)
ggplot() + geom_line(aes(x = x, y = y))


# Example 3

myfun <- function(x, a) {
  1 / (1 + exp(-x + 10) / a)
}

xseq = seq(0, 10, 0.01)
yseq = myfun(xseq, 2)
df = data.frame(xseq, yseq)

ggplot(df) +
  geom_line(mapping = aes(x = xseq, y = yseq), color = "blue") +
  ylim(0, 1) +
  labs(title = "Plot Title", 
       x = "Horizontal label", 
       y = "Vertical label")
```


## Bar plot 
```{r}
# Example 1
data <- data.frame(
  name = c("A", "B", "C", "D", "E") ,  
  value = c(3, 12, 5, 18, 45)
  )

ggplot(data, aes(x = name, y = value)) + 
  geom_bar(stat = "identity")

# Example 2
x <- 0:50
y <- dbinom(x, size = 50, prob = 0.33)
ggplot() + 
  geom_bar(aes(x = x, y = y), 
           stat = "identity", 
           width = 0.2)

```


## Box plot
```{r}
# Take sleep as data set
ggplot(sleep, aes(x = group, y = extra, fill = group)) + 
  geom_boxplot()
```

 
## Histogram
```{r}
data <- data.frame(value = rnorm(100))

ggplot(data) + 
  geom_histogram(aes(x = value), bins = 50)
```


## Smooth line
```{r}

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() +
  geom_smooth()

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() +
  geom_smooth(se = FALSE)
```


## Controlling ggplot elements
 
 
### Color a series 
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy),  color = "red")
```


### Set line type and point shapes
```{r}
p_load(ggpubr)
show_point_shapes()
show_line_types()
```


```{r}

x <- seq(-4,4,0.1)
y <- dnorm(x)

ggplot() + 
  geom_line(aes(x = x, y = y), linetype = 'dashed')
ggplot() + 
  geom_line(aes(x = x, y = y), linetype = 'dotted')

ggplot() + 
  geom_point(aes(x = x, y = y), shape = 0)
ggplot() + 
  geom_point(aes(x = x, y = y), shape = 1)
ggplot() + 
  geom_point(aes(x = x, y = y), shape = 16)
```


### Set labels 
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  labs(color = "Cylinder cont.") +
  labs(x = "New x label") +
  labs(y = "New y label") +
  labs(title = "New Plot", subtitle = "New subtitle") +
  labs(caption = "(Based on data from ...)")
```


### Set shape 
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy), shape = 1) 
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy), shape = 2) 

```


### Set tick marks 
```{r}

ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  scale_y_continuous(breaks=seq(0, 60, 5)) +
  scale_x_continuous(breaks=seq(0, 10, 0.5)) 
```


### Set grid lines
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy))

ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  theme(axis.line = element_line(colour = "black"),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank()) 
```


### Display elements of data seperately
```{r}
# use a third selector
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = class))

ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))

ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, size = class))

ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```


### Add legend
```{r}
x  = seq (0, 10, 0.1)
y1 = x^2
y2 = x^3
df = data.frame(x, y1, y2)
ggplot()+
  geom_line(data=df, aes(y = y1, 
                         x = x,
                         colour = "dfdfdfd"), 
            size = 1 ) +
  geom_line(data=df, aes(y = y2, 
                         x = x,
                         colour = "ahahaha"), 
            size = 1) +
  scale_color_discrete(name = "Functions", 
                       labels = c("y = x^2", "y = x^3"))

```


### Facet grid and wrap - Split the graph in 'facets'
```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_wrap(~ class, nrow = 2)

ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(~ class)

ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(drv ~ cyl)

```


### Variations
```{r}

# Now for the variations 
ggplot(data = mpg)  +
  geom_point(mapping = aes (x = displ, y = hwy))

ggplot(data = mpg)  +
  geom_smooth(mapping = aes (x = displ, y = hwy))

# You can combine the two above graphs easily....
ggplot(data = mpg)  +
  geom_point(mapping = aes (x = displ, y = hwy)) +
  geom_smooth(mapping = aes (x = displ, y = hwy)) 

# Same effect as before, but with parameters more conveniently reorganised
ggplot(data = mpg, mapping = aes (x = displ, y = hwy))  +
  geom_point() +
  geom_smooth()

# You can select data also
ggplot(data = mpg, mapping = aes (x = displ, y = hwy))  +
  geom_point(mapping = aes(color = class)) +
  geom_smooth(
    data = filter(mpg, class == "subcompact"))

ggplot(data = mpg, mapping = aes (x = displ, y = hwy))  +
  geom_point(mapping = aes(color = class)) +
  geom_smooth(data = filter(mpg, class == "subcompact"),
              se = FALSE)

ggplot(data = mpg, mapping = aes (x = displ, y = hwy))  +
  geom_point(mapping = aes(color = class)) +
  geom_smooth( data = filter(mpg, class == "pickup"))

ggplot(data = mpg)  +
  geom_smooth(mapping = aes (x = displ, y = hwy,  class = drv)) +
  geom_hline(yintercept = 20, col = 'green')
```


### Set limits
```{r}
x <- seq(-4, 4, 0.1)
y <- dnorm(x)

ggplot() + 
  geom_line(aes(x = x, y = y)) + 
  ylim(-0.1, 0.7) + 
  xlim(-2, 2)
```


### Horizontal and vertical line
```{r}

n = 10
ggplot(data.frame(x = 1:n, y = 1:n), 
       aes(x = x, y = y)) +
  geom_hline(yintercept = 2, col = 'red') + 
  geom_hline(yintercept = 4, col = 'green') +
  geom_vline(xintercept = 2) + 
  geom_line(size = 1)
```


### Plot in a grid
```{r}

# Needs library gridExtra

p <- ggplot(data = mpg)
p1 <- p + geom_point(mapping = aes(x = displ, y = hwy, color = as.factor(cyl))) 
p2 <- p + geom_point(mapping = aes(x = displ, y = cyl, color = hwy))
    
gridExtra::grid.arrange(
    p1,
    p2,
    nrow = 1)
```


# Wrangling

## Remove NA values
```{r}
# Create a tibble with NA values in it

dd = tibble(x = 1:10,
            y = x^2,
            z = y - 60)
dd[2,2] = NA
dd[2,3] = NA
dd[7,2] = NA
dd[8,2] = NA
dd[6,3] = NA
dd

# Remove all rows where a NA occurs
dd[complete.cases(dd), ]

# Remove all records where a NA occurs in column x
dd1 <- dd[complete.cases(dd['x']), ]

# Remove all records where a NA occurs in column y
dd2 <- dd[complete.cases(dd['y']), ]

dd3 <- dd[complete.cases(dd['z']), ]

# Can do the same with
dd
drop_na(dd,c(y))
drop_na(dd,c(z))
drop_na(dd,c(y,z))
drop_na(dd)
```

## Address elements in an datastructure
```{r}
Input =("
Stream                     Fish
 Mill_Creek_1                76
 Mill_Creek_2               102
 North_Branch_Rock_Creek_1   12
 North_Branch_Rock_Creek_2   39
 Rock_Creek_1                55
 Rock_Creek_2                93
 Rock_Creek_3                98
 Rock_Creek_4                53
 Turkey_Branch              102
")

Data = read.table(textConnection(Input), header=TRUE)
glimpse(Data)

Data[1,1]
Data[1,"Fish"]
Data[1,]$Fish

Data$Stream = as.character(Data$Stream)
Data$Fish = as.numeric(Data$Fish)
glimpse(Data)

colnames(Data)[colnames(Data) == "Fish"] <- "New Fish"

Data <- mutate(Data, log(`New Fish`))

```


## Read a file
```{r}
p_load(devtools)
if (file.exists("/Users/hans/")) {
  setwd("/Users/hans/Documents/R Projects/Learning")
} else {
  setwd("/Users/hansbakker/Documents/R Projects/Learning")
}
getwd()
setwd("..")
getwd()
setwd("Learning")
getwd()

#d = read.csv("./Case Study/Phenobarbital_TH.csv")
#d = read.csv("/Users/hans/Documents/R_Projects/Computational Thinking/Case Study/Phenobarbital_TH.csv")

getwd()
setwd("../Computational Thinking")
getwd()
setwd("./Case Study")
getwd()
if (dir.exists("/Users/hans/")) {
  setwd("/Users/hans/Documents/R_Projects/Computational Thinking/Case Study")
} else {
  setwd("/Users/hansbakker/Documents/R_Projects/Computational Thinking/Case Study") 
}

# requires library readxl
library (readxl)

# lsx1 <- read_excel(path = "Lopinavir_Pregnancy.xlsx",
#                    sheet = "Patient Characteristics",
#                    col_names = T)
# lsx2 <- read_excel(path = "Lopinavir_Pregnancy.xlsx", 
#                    sheet = "Concentrations",
#                    col_names = T, 
#                    skip = 1)
```


## Gather and spread
```{r}
ChickWeight
cw <- ChickWeight %>% as_tibble

cw_w <- cw %>% 
  spread(key = Time, value = weight)
cw_w

cw_l <- cw_w %>% 
  gather(key = "Time", value = "Weight", c(`0`:`21`))
cw_l

# You can arrange the output

cw_l %>%
  arrange(as.numeric(Time), desc(Diet))

# But for some reason you can not arrange on Chick
```


## Detect outlier
```{r}

library(dslabs)

# The data set contains a big outlier
data(outlier_example)
outlier_example

# You can see that because mean and median differ a lot
mean(outlier_example)
median(outlier_example)

# You can see that also because sd is large
sd(outlier_example)

# You can see that also in the histogram
qplot(outlier_example)

# You can see that also in the boxplot
boxplot(outlier_example)

# See if there are values > 8. See! There is one 180 
outlier_example[outlier_example > 8]

# At what index des it sit?
which(outlier_example > 8)

# Remove it form the dataset
oe <- outlier_example[-which(outlier_example>8)]
mean(oe)
median(oe)
sd(oe)
qplot(oe)
boxplot(oe)

```


## Select - columns
```{r}
iris

# For the selection purpose just consider the first few records 
short_iris <- head(iris, 20)

select(short_iris, starts_with("Petal"))
select(short_iris, Species, everything())   #reorder
select(short_iris, -Sepal.Length)
```


## Filter - rows
```{r}
starwars
class(starwars)

# It works on a dataframe

filter(starwars, species == 'Human')
filter(starwars, mass > 100)
filter(starwars, species == 'Human', mass > 100)
filter(starwars, species == 'Human' & mass > 100)
filter(starwars, species == 'Human' | mass > 100)

# It works just as well on a tibble

tib_starwars = as.tibble(starwars)
class(tib_starwars)
filter(tib_starwars, species == 'Human')
```


## Mutate 
```{r}

library(modelr)
library(measurements)

h <- heights %>%
  mutate(
    weight_kg_man = weight * 0.453592,
    weight_kg_pkg = conv_unit(weight, "lbs", "kg"),
    height_m      = conv_unit(height, "inch", "m")
  ) %>%
  select(starts_with('weight'), starts_with("height"))
```

```{r}
my_heights <- heights %>%
  mutate(
    weight = conv_unit(weight, "lbs", "kg"),
    height = conv_unit(height, "inch", "m"),
    bmi    = weight / height  ^ 2
  ) %>% select(bmi, education)
my_heights
```

```{r}
h <- my_heights %>%
  arrange(bmi) %>%
  select(bmi, education)
h
```


## Create data for testing
```{r}

tribble(
  ~colA, ~colB, ~colC,
  "a",   1,     0.1,
  "b",   2,     0.2,
  "c",   3,     0.3
)

tibble(colA = c("a","b","c"),
       colB = c(1,2,3),
       colc = c(0.1, 0.2, 0.3))
```


## Inspect data
```{r}
head(mpg)
glimpse(mpg)
names(mpg)
str(mpg)

# How many values exist? 
count(mpg, model)
count(mpg, trans)

mpg %>% 
  group_by(model) %>%
  summarise(n = n())
```


## Summarise / Group
```{r}
# Summarise two variables
summarise(mpg, 
          mean_hwy = mean(hwy),
          mean_cyl = mean(cyl, na.rm = TRUE))

# Summarise a variable grouped by another variable
s <- group_by(mpg,class)
summarise(s, mean_hwy = mean(hwy))

# Summarise a variable grouped by another variable, using pipes
mpg %>%
  group_by(class) %>%
  summarise(mean_hwy = mean(hwy))

# Summarise a variable grouped by two variable, using pipes
mpg %>%
  group_by(class, trans) %>%
  summarise(mean_hwy = mean(hwy),
            mean_cyl = mean(cyl, na.rm = TRUE))
```


# String processing 
```{r}

tolower("Amsterdam")
toupper("Amsterdam")
sub("-", " ", "Amsterdam-NH-Netherlands")
gsub("-", " ", "Amsterdam-NH-Netherlands")

nchar("Amsterdam")
substr("Amsterdam", 7, 9)
paste(substr("Amsterdam", 1, 1), 
      substr("Amsterdam", 7, 9),
      sep = "")
paste0(substr("Amsterdam", 1, 1), 
      substr("Amsterdam", 7, 9))

str_trim("    Amsterdam     ")

```


# Date processing
```{r}
d2 <- Sys.Date()
format(d2, "%a %d %b")
format(d2, "%A %d %b")
format(d2, "%A %d %B")
format(d2, "%A %d %B %Y")
```



# Rstudio

## Clear console
^L
## Block comment
^-Shift-C
